---
title: "Stat 850 Project Proposal for Team KFF"
format: html
editor: visual
---

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository.

Please ensure that you have set your git name and email address on your computer by **modifying** the following commands and running them in the terminal:

```         
git config --global user.name "Mona Lisa"
git config --global user.email "mona@lisa.com"
```

In the proposal as in the final report, all of your code chunks should be set so that `echo = F`, which prevents the code from being printed in the report. You should also set `message = F, warning = F`, which will reduce the amount of additional output which is present in the report. You can customize the execution of code in quarto by using additional options, some of which are detailed [here](https://quarto.org/docs/computations/execution-options.html).

## Data

### Source

The data we are using in this project comes from \[https://data.wa.gov/api/views/f6w7-q2d2/rows.csv?accessType=DOWNLOAD\].

```{r,echo=FALSE, message=FALSE, warning=FALSE}

# Define the URL of the data you want to load
url <- "https://data.wa.gov/api/views/f6w7-q2d2/rows.csv?accessType=DOWNLOAD"

# Load the data from the URL directly into a data frame
data <- read.csv(url)
```

```{r,echo=FALSE,message=FALSE, warning=FALSE}
summary(data) 
str(data) 
```

```{r, echo=FALSE,message=FALSE, warning=FALSE}  library(dplyr)  Cat_Variable <- c("County","City","State","Make","Model","Electric.Vehicle.Type","Legislative.District","Clean.Alternative.Fuel.Vehicle..CAFV..Eligibility","Electric.Utility")   data <- data %>% mutate_at(Cat_Variable, as.factor)}
```

### Description

This dataset contains information about Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) that have been duly registered as of September 30, 2023 and are currently on record with the Washington State Department of Licensing (DOL).

A Battery Electric Vehicle (BEV) is an all-electric vehicle using one or more batteries to store the electrical energy that powers the motor and is charged by plugging the vehicle in to an electric power source. A Plug-in Hybrid Electric Vehicle (PHEV) is a vehicle that uses one or more batteries to power an electric motor; uses another fuel, such as gasoline or diesel, to power an internal combustion engine or other propulsion source; and is charged by plugging the vehicle in to an electric power source.

It provides insights into the demographic distribution, Make, Model, and geographical prevalence of BEVs and PHEVs, contributing to a better understanding of the electric vehicle landscape in the United State.

Here is a data dictionary showing each variable, the type of variable, units, and range of values.Create a simple data frame

+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Name                                              | Description                                                                                                                                                                                         | Type        | Units   | Range/Values                                                      |
+===================================================+=====================================================================================================================================================================================================+=============+=========+===================================================================+
| VIN (1-10)                                        | The 1st 10 characters of each vehicle's Vehicle Identification Number (VIN).                                                                                                                        | Text        | NA      | NA                                                                |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| County                                            | This is the geographic region of a state that a vehicle's owner is listed to reside within. Vehicles registered in Washington state may be located in other states.                                 | Categorical | NA      | 1 = Adams                                                         |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = Alameda                                                       |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 3 = Alexandria                                                    |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 4 = Allen                                                         |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | .........                                                         |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 179 = Yellowstone                                                 |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| City                                              | The city in which the registered owner resides.                                                                                                                                                     | Categorical | NA      | 1 = Aberdeen                                                      |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = Acme                                                          |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | ........                                                          |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 685 = Zillah                                                      |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| State                                             | This is the geographic region of the country associated with the record. These addresses may be located in other states.                                                                            | Categorical | NA      | 1 = AK                                                            |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = AL                                                            |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | .........                                                         |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 43 = WY                                                           |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Postal Code                                       | The 5 digit zip code in which the registered owner resides.                                                                                                                                         | Numeric     | NA      |                                                                   |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Model Year                                        | The model year of the vehicle, determined by decoding the Vehicle Identification Number (VIN).                                                                                                      | Numeric     | Year    | Min = 1997                                                        |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | Max = 2024                                                        |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Make                                              | The manufacturer of the vehicle, determined by decoding the Vehicle Identification Number (VIN).                                                                                                    | Categorical | NA      | 1 = ALFA ROMEO                                                    |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = AUDI                                                          |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | .......                                                           |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 37 = WHEEGO ELECTRIC CARS                                         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Model                                             | The model of the vehicle, determined by decoding the Vehicle Identification Number (VIN).                                                                                                           | Categorical | NA      | 1= 330E                                                           |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = 500                                                           |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 3 = 530E                                                          |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | ......                                                            |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 127 = XC90                                                        |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Electric Vehicle Type                             | This distinguishes the vehicle as all electric or a plug-in hybrid                                                                                                                                  | Categorical | NA      | 1 = Battery Electric Vehicle (BEV)                                |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = Plug-in Hybrid Electric Vehicle (PHEV)                        |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Clean Alternative Fuel Vehicle (CAFV) Eligibility | This categorizes vehicle as Clean Alternative Fuel Vehicles (CAFVs) based on the fuel requirement and electric-only range requirement in House Bill 2042 as passed in the 2019 legislative session. | Categorical | NA      | 1 = Clean Alternative Fuel Vehicle Eligible                       |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 2 = Eligibility unknown as battery range has not been researched. |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 3 = Not eligible due to low battery range                         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Electric Range                                    | Describes how far a vehicle can travel purely on its electric charge.                                                                                                                               | Numeric     | Miles   | Min =0.00                                                         |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | Max = 337                                                         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Base MSRP                                         | This is the lowest Manufacturer's Suggested Retail Price (MSRP) for any trim level of the model in question.                                                                                        | Numeric     | Dollars | Min = 0.00                                                        |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | Max = 845000                                                      |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Legislative District                              | The specific section of Washington State that the vehicle's owner resides in, as represented in the state legislature.                                                                              | Categorical | NA      | 1 to 49 with 1 representing District 1                            |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | NA indicates missing data                                         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| DOL Vehicle ID                                    | Unique number assigned to each vehicle by Department of Licensing for identification purposes.                                                                                                      | Numeric     | NA      | Min = 4385                                                        |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | Max = 479254772                                                   |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Vehicle Location                                  | The center of the ZIP Code for the registered vehicle.                                                                                                                                              | Character   | NA      |                                                                   |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| Electric Utility                                  | Data of Washington vehicle addresses and electric power retail service territories, classified by type and combined using geographic information systems                                            | Categorical | NA      | 1 =Avista Corp                                                    |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | .......                                                           |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | 76 = PUGET SOUND ENERGY INC\|\|PUD NO 1 OF WHATCOM COUNTY         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
| 2022 Census Tract                                 | Combination of the state, county, and census tract codes as assigned by the United States Census Bureau in the 2020 census, also known as Geographic Identifier (GEOID).                            | Numeric     | NA      | Min = 1.081e+09                                                   |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | Max = 5.603e+10                                                   |
|                                                   |                                                                                                                                                                                                     |             |         |                                                                   |
|                                                   |                                                                                                                                                                                                     |             |         | NA indicates missing data                                         |
+---------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+---------+-------------------------------------------------------------------+
